\documentclass{article}

\usepackage{extramarks}
\usepackage{amsmath}
\usepackage{amsthm}
\usepackage{tikz}
\usepackage{enumerate}
\usepackage{comment}

\usetikzlibrary{automata, positioning}

\topmargin=-0.4in
\evensidemargin=0in
\oddsidemargin=0in
\textwidth=6.5in
\textheight=9.0in

\linespread{1.1}

\newcommand{\class}{CS244 Theory of Computation}
\newcommand{\authorName}{Yuyang(Peter) Rong}

\title{
    \vspace{-0.8in}
    \textmd{\textbf{\class \\ Cheatsheet - Midterm}}\\
}

\author{\authorName}
\date{}


\newcommand{\TM}{\mathsf{TM}}
\newcommand{\ATM}{A_\TM}
\newcommand{\TR}{Turing-recognizable}
\newcommand{\De}{Decidable}
\newcommand{\U}{\Sigma^*}
\newcommand{\lem}{\leq_\mathrm{m}}

\begin{document}

\maketitle

\begin{comment}
\setcounter{section}{-1}
\section{Introduction}\begin{enumerate}
	\item Proof ideas are important? Too lazy to list them out...
\end{enumerate}
\end{comment}


\section{Regular languages \& finite automata}\begin{enumerate}
	\item $DFA \equiv NFA \equiv \epsilon-NFA \equiv RE \equiv RLG$
	\item Pumping Lemma. $L$ is $RE$, $\exists p > 1, \forall s \in L \text{where} |s| \ge p, s = xyz$:
	\begin{enumerate}
		\item $\forall i \ge 0, xy^iz \in L$
		\item $|y| \ge 1$
		\item $|xy| \le p$
	\end{enumerate}
	Take special notes on
	\begin{enumerate}
		\item Sometimes it can be \textbf{equal}
		\item Satisfying this lemma \textbf{does not} mean the the language is RE.
	\end{enumerate} 
	\item Myhill-Nerode Theorem. $\forall u, v\in \U$, $u \sim_L v \Longleftrightarrow (\forall z \in \U, uz \in L \Leftrightarrow vz \in L)$ \\
	$$L \text{ is regular} \Longleftrightarrow \sim_L \text{ is finite.} $$
	\item Closure properties: closed under $L_1 \cup L_2$, $L_1L_2$, $L^*$
\end{enumerate}


\section{Context free language \& push down automata}\begin{enumerate}
	\item Chomsky normal form. Everyrule is either $A \rightarrow BC$ or $A \rightarrow a$, only $S$ can be $\epsilon$
	\item Pumpling lemma. $L$ is $CFG$, $\exists p > 1, \forall s \in L \text{where} |s| \ge p, s = uvwxy$:
	\begin{enumerate}
		\item $\forall i \ge 0, uv^iwx^iy \in L$
		\item $|vx| \ge 1$
		\item $|vwx| \le p$
	\end{enumerate}
	Take special notes on
	\begin{enumerate}
		\item Sometimes it can be \textbf{equal}
		\item Satisfying this lemma \textbf{does not} mean the the language is RE.
	\end{enumerate}
	\item Closure properities:
	\begin{enumerate}
		\item $L_1 \cup L_2$, $L \cap RE$ closed
		\item $L_1 \cap L_2$, $\overline{L}$ not closed
	\end{enumerate}
	\item Deterministic context free language(DCFL) \& deterministic push down automata(DPDA)
	\begin{enumerate}
		\item DCFL is closed under complementation; any CFL whose complement is not CFL is not DCFL
		\item $A$ is DCFL $\Longleftrightarrow$ $A\dashv$ is DCFL
	\end{enumerate}
\end{enumerate}


\section{Turing machine}\begin{enumerate}
	\item Multitape $\TM$ has an equivalent single-table $\TM$. So is non-deterministic $\TM$
	\item $A$ is \TR $\Longleftrightarrow \exists$ Enumerator $E$ enumerates it.
\end{enumerate}


\section{Decidability}\begin{enumerate}
	\item $A_{CFG}$ \De, simulate for $2|w| - 1$ times.
	\item $\ATM$ is not \De but \TR, $\overline{\ATM}$ is not \TR, proof of $\ATM$ being not \De
	\begin{enumerate}
		\item Diag method. $H$ decides $\ATM$, construct $D$ on $\langle M \rangle$:
		\begin{enumerate}
			\item Run $H$ on $\langle M, \langle m \rangle \rangle$
			\item Output the opposite of H.
		\end{enumerate}
		\item $\ATM \lem HALT_{\TM}$
		\item Using Recursion Theorm, construct $B$ on input $w$:
		\begin{enumerate}
			\item Obtain $\langle B \rangle$
			\item Run the decider $R$ of $\ATM$ on $\langle B, w \rangle$
			\item Output the opposit of R.
		\end{enumerate}
	\end{enumerate}
	\item $A$ \De $\Longleftrightarrow$ $A$ is \TR and $\overline{A}$ is \TR
	\item $\overline{\ATM} \text{ is not \TR}$
\end{enumerate}


\section{Reducibility}\begin{enumerate}
	\item Reduce to $\ATM$: If this is decidable, then $\ATM$ is too.
	\item Via Computation histories
	\item Mapping Reducibility: $f: \U \rightarrow \U, \forall w \in \U$, $w \in A \Longleftrightarrow f(w) \in B$
	\begin{enumerate}
		\item $A \lem B$.  \\
			 $B$ is \De $\rightarrow A$ is \De;  \\
			$A$ is not \De $\rightarrow B$ is not \De.
		\item $A \lem B$.  \\
			 $B$ is \TR $\rightarrow A$ is \TR;  \\
			$A$ is not \TR $\rightarrow B$ is not \TR.
	\end{enumerate}
\end{enumerate}


\section{Advanced topics in computability theory}\begin{enumerate}
	\item $\langle SELF \rangle$
	\item Turing Reducibility
\end{enumerate}

\end{document}